---
title: "Predict publishing date"
author: "Carsten Behring"
date: "3 Feb 2015"
output: html_document
---

```{r imports, results="hide"}
library(dplyr)
library(stringr)
library(knitr)
library(pander)
library(ggplot2)
library(kernlab)
library(caret)
library(doMC)
registerDoMC(4)
set.seed(1234)
```
I just read into R the complete export of the register of questions.
 
```{r read data, cache=T}
questions <- read.csv("Questions_update20150914134554347.csv",encoding="UTF-8")
questions$QUESTIONTYPE <-  iconv(questions$QUESTIONTYPE,"UTF-8","ASCII",sub="")

questions <- questions %>%
    mutate(RECEPTIONDATE = as.Date(RECEPTIONDATE,format="%d/%m/%Y")) %>%
    mutate(PUBLISHINGDATE = as.Date(PUBLISHINGDATE,format="%d/%m/%Y")) %>%
    mutate(reception.dayoffset=as.numeric(RECEPTIONDATE - as.Date("2000-01-01")))
```


As I want to predict the publication date from (potentially) all other variables, lets see how many non empty publication dates we have, to see the maximum of obsevations usable for training.

There are `r sum(is.na(questions$PUBLISHINGDATE))` observations with empty publishing date, which leaves `r nrow(questions) - sum(is.na(questions$PUBLISHINGDATE))` usefull observations.


```{r}
observations <- questions %>% filter(!is.na(questions$PUBLISHINGDATE))
```

## Feature selection
Which features from the dataset should I use to predict the publishing date.
The dataset contains the following columns:

```{r}
 pander(data.frame(names(observations)))
```


From intuition, I select the following:

* UNIT
* Applicant
* Reception Date
* Deadline
* Foodsectorarea
* Questiontype
* Mandatetype
* MandateRequestor


Out of these, the recption date and acceptance date (and the publishing date), I need to recode.
Leaving them as dates, gives probbaly too little observations 'for each date', so I will recode them as 'months' .

And in reality I dont't want to predict a publishing date, but how long (= how many month ) did the question have between receptiondate date and publishing date.

Lets see first, for how many observations I can calculate this 'duration', so which have both 'reception date' and 'publication date' set.


```{r}
 observations <- observations %>% filter( !is.na(observations$RECEPTIONDATE))
```

We have `r nrow(observations)` where a duration can be calculated.



An other important feature is the clock stop information. This comes from an other database extract and gets joined with the existing data.

```{r}

clock_stops<-read.csv("Questions_clock_update20150915110216567.csv",stringsAsFactors=F) %>%
    select( Question.Number,Clock.stop.date,Clock.start.date,Difference..EFSA.working.days.) %>%
    mutate(Clock.stop.date = as.Date(Clock.stop.date,format="%d/%m/%Y")) %>%
    mutate(Clock.start.date = as.Date(Clock.start.date,format="%d/%m/%Y")) %>%
    mutate(Difference..EFSA.working.days. = as.numeric(Difference..EFSA.working.days.)) %>%
    filter(!is.na(Difference..EFSA.working.days.))

clock_stops_counts <- clock_stops %>%
    group_by(Question.Number) %>%
    mutate(count=n()) %>%
    mutate(total.stopp = sum(Clock.start.date - Clock.stop.date)) %>%
    mutate(total.stop.wd = sum(Difference..EFSA.working.days.)) %>%
    rename(QUESTIONNUMBER=Question.Number) %>%
    select(QUESTIONNUMBER,count,total.stopp,total.stop.wd)
    
```


```{r}
durations <- as.numeric(observations$PUBLISHINGDATE -observations$RECEPTIONDATE)

observations <- observations %>%
    mutate(duration= round((as.numeric(PUBLISHINGDATE -  RECEPTIONDATE)) / 30))


observations <- left_join(observations,clock_stops_counts) %>%
    rename(clock.stop.count=count)
   
    

observations <- observations %>%
    filter(duration > 0) %>%
    mutate(duration.net = as.numeric(round(duration-(total.stopp/30)))) %>%
    mutate(duration.net = ifelse(is.na(duration.net),duration,duration.net)) %>%
    filter(duration.net > 0) %>%
    mutate(clock.stop.count = ifelse(is.na(clock.stop.count),0,clock.stop.count))
    


ggplot(observations ,aes(duration)) + geom_bar(binwidth=1)
ggplot(observations ,aes(duration.net)) + geom_bar(binwidth=1)

```

This show the distribution of the 'durations' in months. The majority is between 1 and 40 months and most durations have several observations, which should allow the training to work reasonably.


The following plots show the relationship of all considered variables with the duration.

```{r}

qplot(x=duration,y=UNIT,data=observations)  + geom_jitter()
qplot(x=duration,y=APPLICANT,data=observations) + geom_jitter()
qplot(x=duration,y=FOODSECTORAREA,data=observations) + geom_jitter()
qplot(x=duration,y=QUESTIONTYPE,data=observations) + geom_jitter()
qplot(x=duration,y=MANDATETYPE,data=observations) + geom_jitter()
qplot(x=duration,y=MANDATEREQUESTOR,data=observations) + geom_jitter()
qplot(x=duration,y=clock.stop.count,data=observations) + geom_jitter()
qplot(x=duration,y=reception.dayoffset,data=observations) + geom_jitter()
qplot(x=duration,y=months(observations$RECEPTIONDATE),data=observations) + geom_jitter()
```

From the plots, there are very view obvious correlations between any single variable with the duration. These are:

- UNIT: Pesticides, Nutrition, GMO, FIP units have longer durations then the others
- APPLICANT: DG Sanco have longest duration and an other group (UK,NL,FR,EFSA) have a longer duration as the rest
- FOODSECTORAREA: Dietic products and Flavourings have laonger durations then others
- clockstop.count: Clockstops result in a loger minimal duration


I use random forest model to create the first model

```{r cache=T}
observations.train <- sample_n(observations,1000) 
inTrain <- createDataPartition(y = observations.train$duration,p=0.75,list=F)
training <- observations.train[inTrain,]
testing <- observations.train[-inTrain,]
fit <- train(duration ~ UNIT + APPLICANT + FOODSECTORAREA + QUESTIONTYPE+ MANDATETYPE + MANDATEREQUESTOR 
            + clock.stop.count + reception.dayoffset + months(RECEPTIONDATE), data = training,method = "rf",importance=T)

```

The following plot shows the importance of the top most important variables.

```{r}
 imp <- varImp(fit)
 plot(imp,top=20)
```

Result of training:

```{r}
fit

fit$finalModel

```

Residual plot:

```{r}
testing <- testing[testing$QUESTIONTYPE %in% training$QUESTIONTYPE,]
pred <-  predict(fit,newdata = testing)

residuals <- pred - testing$duration

plot(residuals)
hist(residuals)

```

SME for testing:

```{r}
 summary(abs(residuals))
 R2(pred,testing$duration)
RMSE(pred,testing$duration)
cor(pred,testing$duration)
```

The distribution of the absolute residuals

```{r}
boxplot(abs(residuals))
```

shows that the mean error of the model is at +/-
`r mean(abs(residuals))` months.
